---
title: "Problem Set 3"
author: "Jay Hesselberth"
date: "February 20, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview 

We will examine a [time-course of mRNA expression][1] of barley powdery mildew fungus in immunocompromised plants (*Arabidopsis*). The fungus Blumeria graminis f. sp. hordei (called *Bgh* from now on)

Counts were generated using the [`maSigPro` software][3] and deposited in [NCBI GEO][2].

[1]: http://www.ncbi.nlm.nih.gov/pubmed/23175786
[2]: http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE43163  
[3]: http://bioinformatics.oxfordjournals.org/content/30/18/2598.full

Some important identifiers for the data set:

| Name | Description |
| ---- | ----------- |
| `A6` | *Bgh* isolate expressing other AVRA effectors |
| `K1` | *Bgh* isolate expressing the cognate AVRA1 effector for MLA1|
| `pps` | *Arabidopsis* plants |
| `B12` | *Arabidopsis* plants expressing MLA1-HA|

We will download tables from GEO containing counts of mRNA abundance from both the fungal sample.

# Raw data

First we need load the data.

```{r load_raw_data, message = FALSE}
library(readr)
library(tidyr)
library(dplyr)
library(stringr)

# bgh data
bgh_url <- 'http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE43163&format=file&file=GSE43163_CompleteCountTable_Bgh.txt.gz'

# 2 info and blank lines at top, skip them
raw_tab_bgh <- read_tsv(bgh_url, skip = 2)

# the header for the first column is "NA", set it manually
names(raw_tab_bgh)[1] <- 'gene.name'

head(raw_tab_bgh)
```

# Problems

## Problem 1

Now that the raw data are loaded, your first task is to tidy the data with `tidyr`.

```{r problem_1}
# add tidying code here
cleaned_data <- raw_data %>% tbl_df
cleaned_data
```

## Problem 2

```{r problem_2}
```

## Problem 3

```{r problem_3}
```

## Problem 4

```{r problem_4}
```

## Problem 5

```{r problem_5}
```
