---
title: "Problem Set 3. Due Thurs March 2 5pm"
author: "Sarah Stonedahl"
date: "Feb 23, 2017"
output: html_document
---
URL: http://rpubs.com/stonedas/256869

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
```

```{r}

```
# Submission

1. Fork this repository to your own account
1. Make changes to the `README.Rmd` file (including the author field above).
1. Knit the file to HTML format and add it to your git repository (`git add`)
1. Submit a pull request with your Rmd and knitted HTML files.

# Overview 

You will examine a [time-course of mRNA expression][1] of barley powdery mildew fungus in immunocompromised plants (*Arabidopsis*). The fungus Blumeria graminis f. sp. hordei (called *Bgh* from now on)

Counts were generated using the [`maSigPro` software][3] and deposited in [NCBI GEO][2].

[1]: http://www.ncbi.nlm.nih.gov/pubmed/23175786
[2]: http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE43163  
[3]: http://bioinformatics.oxfordjournals.org/content/30/18/2598.full

Some important identifiers for the data set:

| Name | Description |
| ---- | ----------- |
| `A6` | *Bgh* isolate expressing other AVRA effectors |
| `K1` | *Bgh* isolate expressing the cognate AVRA1 effector for MLA1|
| `pps` | *Arabidopsis* plants |
| `B12` | *Arabidopsis* plants expressing MLA1-HA|

We will download tables from GEO containing counts of mRNA abundance from both the fungal sample.

# Raw data

First we need load the data.

```{r load_raw_data, message = FALSE}
library(tidyverse)

# bgh data
bgh_url <- 'http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE43163&format=file&file=GSE43163_CompleteCountTable_Bgh.txt.gz'

# 2 info and blank lines at top, skip them
raw_data <- read_tsv(bgh_url, skip = 2)

# the header for the first column is "NA", set it manually
names(raw_data)[1] <- 'gene.name'

raw_data <- as_data_frame(raw_data)
```

# Problems


## Problem 1

Now that the raw data are loaded, your first task is to tidy the data with `tidyr`. The `cleaned_data` should look something like this:

```
# A tibble: 310,896 × 6
             gene.name plant fungus time.point   rep value
*                <chr> <chr>  <chr>      <chr> <chr> <int>
1             bgh04079   B12     A6       6hpi     1    13
2             bgh01634   B12     A6       6hpi     1    31
3  bghG000012000001001   B12     A6       6hpi     1   121
4  bghG000012000002001   B12     A6       6hpi     1     3
5             bgh00757   B12     A6       6hpi     1   253
6             bgh01273   B12     A6       6hpi     1    45
7             bgh01274   B12     A6       6hpi     1    32
8             bgh01277   B12     A6       6hpi     1     9
9             bgh06140   B12     A6       6hpi     1    47
10            bgh05774   B12     A6       6hpi     1    42
# ... with 310,886 more rows
```

The key steps are to `gather()` the data into key-value pairs and then `separate()` the information from the key column into new columns.

```{r problem_1}
raw_data %>% gather(key, value, -gene.name) %>% separate (key, into = c ('plant', 'fungus', 'time.point', 'rep'), sep = '_') -> cleaned_data

 # report the cleaned_data by just naming it, uncomment the following line:
cleaned_data

```

## Problem 2
You need to translate what you learned on the command line into the R world.

Which plant has the highest expression of any gene in the 6hpi time point?

```{bash, eval=FALSE}
raw_data | awk '$4 == "6hpi"' | sort -k6nr | head -n 1 | cut -f2
```

```{r}
library(stringr)
cleaned_data %>% mutate(time.point = str_replace(time.point, 'hpi', ''), time.point = str_c('hpi.', time.point)) %>% filter(time.point == "hpi.6") %>% arrange(desc(value)) %>% slice(1:1) %>% select(plant)
```

Which plant / fungus pair has the highest expression in the 18hpi time point?

```{bash, eval=FALSE}
raw_data | awk '$4 == "18hpi"' | sort -k6nr | head -n 1 | cut -f2,3
```

```{r}
cleaned_data %>% mutate(time.point = str_replace(time.point, 'hpi', ''), time.point = str_c('hpi.', time.point)) %>% filter(time.point == "hpi.18") %>% arrange(desc(value)) %>% slice(1:1) %>% select(plant)
```

## Problem 3

Identify the top 3 most consistently differentially expressed genes between the earliest and latest time points for each combination of `plant` and `fungus` strains.

- "Differential expression"" is the difference between `value` (i.e., gene expression level) between time points.

- "Consistency" is the smallest variance in `value` between replicates.

### Strategy

1. Create a new table from the cleaned data by moving each `hpi` value to a new column name, with counts for each in the column (hint: use a `tidyr` verb). It is helpful to reformat the `hpi` values by converting from e.g. `6hpi` to `hpi.6`. You can use `mutate` to do this, i.e.:

```{r reformat_time, eval=FALSE}

```

2. Create a new column containing the expression difference between the relevant time points.

3. Calculate summary statistics (mean and variance) of the expression differences by grouping (hint) the `gene.name`, `plant`, and `virus` columns.

4. Sort by these statistics and use the `dplyr` verb `slice` to pull the ones you want (i.e., the top 3). Note you will have to remove `gene.name` from the grouping so that sorting works.

```{r problem_3}
cleaned_data %>% mutate(time.point = str_replace(time.point, 'hpi', ''), time.point = str_c('hpi.', time.point)) %>% spread(time.point, value) -> cleaned_data_hpi

cleaned_data_hpi %>% mutate(expr.diff = hpi.24 - hpi.6) -> expr_data

expr_data %>% group_by(gene.name, plant, fungus) %>% summarise(mean_expr = mean(expr.diff), var_exp = var(expr.diff)) %>% ungroup(gene.name) %>% arrange(desc(mean_expr), var_exp) %>% group_by(plant, fungus) %>% slice(1:3)

expr_data %>% group_by(gene.name, plant, fungus) %>% summarise(mean_expr = mean(expr.diff), var_exp = var(expr.diff)) %>% ungroup(gene.name) %>% arrange(desc(mean_expr), var_exp) %>% group_by(plant, fungus) %>% slice(1:3) -> con_diff_data

```

## Problem 4

Now examine the above final data frame above and write a few sentences putting inline code in least 3 places. For example, There are `r nrow(mtcars)` rows of data in `mtcars`.

There are `r nrow(con_diff_data)` rows and `r ncol(con_diff_data)` columns in `con_diff_data`. There are `r n_groups(con_diff_data)` combinations of plant and fungus in `con_diff_data`.


## Problem 5

Plot the expression (`value`) by time (`hpi`) for the above genes. Format the plot as follows:

1. Plot each point.
1. Connect the points with a smooth line 
1. Create a facet for each `plant` / `bgh` pair 
1. Assign a different color to each gene 
1. Assign a different shape to each replicate 
1. Add a descriptive title to the plot 
1. Use the "black & white" theme

Extra credit: add error bars to the plot (use `geom_errorbar`).

```{r problem_5}
semi_join(cleaned_data, con_diff_data, by = "gene.name") -> joined_data

joined_data %>% ggplot(aes(x = time.point, y = value, color = gene.name, shape = rep)) + geom_point() + facet_grid(fungus ~ plant) + geom_smooth(method = "lm", se = FALSE) + theme_bw() + labs(title = "Gene expression over time")
```
